From Reversible Programs to \\ Univalent Universes and Back

### 2021-09-05 Pi+/paper/discussion.tex

Higher-order term-rewriting systems and word problems have a long history of being formalised in proof assistants like
{homotopy.io}, Agda, Coq and Lean~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main
result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem. \citet{Hiver-coq}
describes an explicit algorithm for producing normal forms. could provide an alternative to our rewriting system. Other
encodings of permutations as listed vectors, matrices, inductively generated trees (Motzkin trees), Young diagrams, or
string diagrams, proved either difficult formalise in type theory or difficult to relate directly to the primitive type
isomorphisms of $\PiLang$. The automatic \citet{knuthSimpleWordProblems1970} completion produced too many equations
making proving correctness and termination intractable.

Coherence theorems are famous problems in category theory, and Mac Lane's coherence
theorem~\cite{maclaneNaturalAssociativityCommutativity1963,joyalBraidedTensorCategories1993,gurskiInfiniteLoopSpaces2013}
for monoidal categories is a particular one. The use of rewriting and proof assistants to prove coherence theorems
for higher categorical structures has a long history
see~\cite*{forestCoherenceGrayCategories2018,beylinExtractingProofCoherence1996}.

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule
$\term{ua_\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule
$\term{ua_\eta} : (p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when
formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full univalence axiom
$(A \eqv B) \eqv (A \id_{\UU} B)$.
% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction. The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which,
in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean
circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the existence
of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact
that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
provides an intuition on why certain constructions are hard (or impossible in the general case). There are other,
different approaches to computing with univalence,
in~\cite*{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in
Cubical Type
Theory\cite*{angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.

>>>>

Higher-order term-rewriting systems and word problems have a long history of being formalised in proof assistants like
{homotopy.io}, Agda, Coq and Lean~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main
result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem. \citet{Hiver-coq}
describes an explicit algorithm for producing normal forms. It could provide an alternative to our rewriting system. Other
encodings of permutations as listed vectors, matrices, inductively generated trees (Motzkin trees), Young diagrams, or
string diagrams, proved either difficult to formalise in type theory or difficult to relate directly to the primitive type
isomorphisms of $\PiLang$. The automatic \citet{knuthSimpleWordProblems1970} completion produced too many equations
making proving correctness and termination intractable.

Coherence theorems are famous problems in category theory, and Mac Lane's coherence
theorem~\cite{maclaneNaturalAssociativityCommutativity1963,joyalBraidedTensorCategories1993,gurskiInfiniteLoopSpaces2013}
for monoidal categories is a particular one. The use of rewriting and proof assistants to prove coherence theorems
for higher categorical structures has a long history, see~\cite*{forestCoherenceGrayCategories2018,beylinExtractingProofCoherence1996}.

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule
$\term{ua_\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule
$\term{ua_\eta} : (p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when
formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full univalence axiom
$(A \eqv B) \eqv (A \id_{\UU} B)$.
% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction. The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which,
in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean
circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the existence
of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact
that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
2-combinators). Thus, our results suggest a new computational interpretation of the univalence principle, which
provides an intuition on why certain constructions are hard (or impossible in the general case). There are other,
different approaches to computing with univalence,
in~\cite*{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in
Cubical Type
Theory\cite*{angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.

### 2021-09-05 Pi+/paper/ufin.tex

Functions between types are functors between groupoids, and type families (or functions to the universe) are indexed
families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to A$
which has a lifting operation giving it the structure of a fibration. The $\term{transport}$ operation lifts paths in
the base space to functions between the fibers.  Using the groupoid structure of $A$, for a path $p : x \id_{A} y$,
$\term{transport}(p)$ and $\term{transport}(\inv{p})$ form an equivalence.
% We name this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
\[
  \tptEqv{P} : \dfun*{x,y:A}{(x \id_{A} y) \to (P(x) \eqv P(y))}
\]

>>>>

Functions between types are functors between groupoids, and type families (or functions to the universe) are indexed
families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to A$
which has a lifting operation giving it the structure of a fibration. The $\term{transport}$ operation lifts paths in
the base space to functions between the fibers.  Using the groupoid structure of $A$, for any $x, y : A$ and a path 
$p : x \id_{A} y$, $\term{transport}(p)$ and $\term{transport}(\inv{p})$ form an equivalence.
% We name this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
\[
  \tptEqv{P} : (x \id_{A} y) \to (P(x) \eqv P(y))
\]
